Phonetic Tansfer analysis code

### 2022-10-01 RCode/transfer_analysis.Rmd
trial commit; 2nd round revisions

options(scipen=999) # turn off scientific e notation, show full numbers. To turn on: options(scipen=0)
```

@@

```{r}
## final result plot:

@@

# raw FFs
with(df_lobanov_gender_time, plotVowels(f1, f2, vowel, group = gender,
                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
                               var.col.by = vowel, var.sty.by = gender, 
                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
                               ellipse.line = T,  ellipse.fill = T,
                               poly.line = F, poly.order = c("\\ae","\\vt"),
                               xlim = xlim, ylim = ylim, 
                               legend.kwd = "bottomright", legend.args = gender,
                               fill.opacity = 0.05, pretty = TRUE))

@@

@@

# full model
phon_acco.model_f1 = lmer(f1 ~ 
                         time*vowel + vowel*context*task + time*task +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

@@

# null model
null.model = lmer(f1~
                   time*vowel + vowel*context*task + time*task +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

@@

@@

```{r, eval=FALSE, include=FALSE}
# scratch pad: additional code for manuscript revision

@@

>>>>

options(scipen=0) # turn off scientific e notation: options(scipen=999), show full numbers. To turn on: options(scipen=0)
```

@@

```{r}
## final result plot (unilingual IE + mixed + baseline Bengali):

@@

with(df_lobanov_gender_time, plotVowels(f1, f2, vowel, group = gender,
                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
                               var.col.by = vowel, var.sty.by = gender, 
                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
                               ellipse.line = T,  ellipse.fill = T,
                               poly.line = F, poly.order = c("\\ae","\\vt"),
                               xlim = xlim, ylim = ylim, 
                               legend.kwd = "bottomright", legend.args = gender,
                               fill.opacity = 0.05, pretty = TRUE))

@@

---

### 29.7.2022 (second round revision):

gender normalizing the raw (non-z-score normalized) data

gender_normalized <- with(df_time, normLobanov(cbind(f1,f2), group=gender))

df_gender_time <- df_time %>%
  select(!c(f1,f2)) %>%
  mutate(f1 = gender_normalized$f1,
         f2 = gender_normalized$f2)

rm(gender_normalized)

# Plots before gender normalization:

xlim = c(2500, 700)
ylim = c(1400, 100)

with(df_time, plotVowels(f1, f2, vowel, group = gender,
                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
                               var.col.by = vowel, var.sty.by = gender, 
                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
                               ellipse.line = T,  ellipse.fill = T,
                               poly.line = F, poly.order = c("\\ae","\\vt"),
                               xlim = xlim, ylim = ylim, 
                               legend.kwd = "bottomright", legend.args = gender,
                               fill.opacity = 0.05, pretty = TRUE))

# Plots after gender normalization:

xlim = c(3, -3)
ylim = c(4, -4)

with(df_gender_time, plotVowels(f1, f2, vowel, group = gender,
                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
                               var.col.by = vowel, var.sty.by = gender, 
                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
                               ellipse.line = T,  ellipse.fill = T,
                               poly.line = F, poly.order = c("\\ae","\\vt"),
                               xlim = xlim, ylim = ylim, 
                               legend.kwd = "bottomright", legend.args = gender,
                               fill.opacity = 0.05, pretty = TRUE))

---

@@

# full model
phon_acco.model_f1 = lmer(f1 ~ 
                         time*vowel + vowel*context +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

@@

# null model
null.model = lmer(f1~
                   time*vowel +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

@@

---

### 29.7.2022 (second round revision):

Fitting models to non-z-score-normalized data (only gender normalized)

temp_raw <- df_gender_time %>%
  mutate(time = as.numeric(time)) %>%
  mutate(time = scale(time))

# full model
raw_phon_acco.model_f1 = lmer(f1 ~ 
                         time*vowel + task*vowel*context +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f1)

# null model
raw_null.model = lmer(f1~
                   time*vowel + vowel*context +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f1, raw_null.model)

@@

# scratch pad: additional code for manuscript revision

@@

## 29.7.2022 (second round revision):

Content for separate .rmd file:

Fitting models to non-z-score-normalized data (only gender normalized)

temp_raw <- df_gender_time %>%
  mutate(time = as.numeric(time)) %>%
  mutate(time = scale(time))

### F1

First, to test our first research question (L2 vowels), we fitted a model with vowel\*time, and vowel\*context as fixed effects. Comparing this to a null model that lacks the transfer term shows that the full model is significantly better. To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction. The full model was better.

# full model
raw_phon_acco.model_f1 = lmer(f1 ~ 
                         time*vowel + vowel*context +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f1)

# null model 1
raw_null.model = lmer(f1~
                   time*vowel +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f1, raw_null.model)

# null model 2
raw_null.model = lmer(f1~
                   time*vowel + context +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f1, raw_null.model)

To test the hypothesis Paradigm, we fitted a model with vowel*time and task*vowel*context as fixed effects. This three-way interaction term is significant. Comparing this model to a null model that lacks the fixed effect Task shows that the full model is better. Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term
to test whether the interaction term is needed. However, the full model was still a better fit for the data:

# full model
raw_phon_acco.model_f1 = lmer(f1 ~ 
                         time*vowel + task*vowel*context +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f1)

# null model 1
raw_null.model = lmer(f1~
                   time*vowel + vowel*context +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f1, raw_null.model)

# null model 2
raw_null.model = lmer(f1~
                   time*vowel + vowel*context + task +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f1, raw_null.model)

Because an independent effect of task on FFs is unexpected (c.f. model interpretation below), we wanted to examine if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant and improves model fit.

# full model
raw_phon_acco.model_f1 = lmer(f1 ~ 
                         time*vowel + task*vowel*context + task*time +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f1)

# null model 
raw_null.model = lmer(f1~
                   time*vowel + task*vowel*context +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f1, raw_null.model)

Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel\*context) and time. This term has a significant effect, but does not significantly improve model fit compared to a null model lacking the vowel\*context\*time term.

# full model
raw_phon_acco.model_f1 = lmer(f1 ~ 
                         time*vowel + task*vowel*context + task*time + vowel*context*time +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f1)

# null model 
raw_null.model = lmer(f1~
                   time*vowel + task*vowel*context + task*time +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f1, raw_null.model)

Thus, the optimal model was: f1 ~ time\*vowel + vowel\*context\*task + time\*task + (1 + context|word) + (1 + context + vowel|subject).

### F2

First, to test our first research question (L2 vowels), we fitted a model with vowel\*time, and vowel\*context as fixed effects. Comparing this to a null model that lacks the transfer term shows that the full model is significantly better. To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction. The full model was better.

# full model
raw_phon_acco.model_f2 = lmer(f2 ~ 
                         time*vowel + vowel*context +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f2)

# null model 1
raw_null.model = lmer(f2~
                   time*vowel +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f2, raw_null.model)

# null model 2
raw_null.model = lmer(f2~
                   time*vowel + context +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f2, raw_null.model)

To test the hypothesis Paradigm, we fitted a model with vowel\*time and task\*vowel\*context as fixed effects. While task has an independent effect on F2, the three-way interaction term is not significant. However, there are significant two-way interactions between task\*context and task\*vowel. Comparing this to a null model
that lacks the fixed effect Task confirms that the full model is still a better fit for the data, indicating that an optimal model should contain the term Task. To confirm whether Task interacts with any of the other terms, we fitted a model with Task as a separate fixed effect without the interaction term. Comparing this to the full (interaction) model shows that the latter is still a better fit.

# full model
raw_phon_acco.model_f2 = lmer(f2 ~ 
                         time*vowel + task*vowel*context +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f2)

# null model 1
raw_null.model = lmer(f2~
                   time*vowel + vowel*context +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f2, raw_null.model)

# null model 2
raw_null.model = lmer(f2~
                   time*vowel + vowel*context + task +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f2, raw_null.model)

Because an independent effect of task on FFs is unexpected (c.f. model interpretation below), we wanted to examine if the effect might be local, i.e. decrease over time. We added a task\*time interaction term to the full model, which is significant and improves model fit.

# full model
raw_phon_acco.model_f2 = lmer(f2 ~ 
                         time*vowel + task*vowel*context + task*time +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f2)

# null model
raw_null.model = lmer(f2~
                   time*vowel + task*vowel*context +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f2, raw_null.model)

Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel\*context) and time. This term is not significant, and does not significantly improve model fit compared to a null model lacking the vowel\*context\*time term.

# full model
raw_phon_acco.model_f2 = lmer(f2 ~ 
                         time*vowel + task*vowel*context + task*time + vowel*context*time +
                         (1+context|word) + (1+context+vowel|subject), 
                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_phon_acco.model_f2)

# null model
raw_null.model = lmer(f2~
                   time*vowel + task*vowel*context + task*time +
                   (1+context|word) + (1+context+vowel|subject), 
                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))

summary(raw_null.model)

anova(raw_phon_acco.model_f2, raw_null.model)

Thus, the optimal model is: f2 ~ vowel\*time + vowel\*context\*task + task\*time + (1+context|word) + (1+context+vowel|subject).
